---
title: "Data_Wrangling_Visualization_In_Class_Exercises"
output: pdf_document
date: "2026-08"
---

1) Download the file *In_Class_Exercises_Arsenic.csv* file associated with this week's in class assignment. This data is adapted from the 2015-2016 National Health and Nutrition Examination Survey (NHANES). We've truncated the data to include only a few demographic variables along with urinary arsenic levels. Using base R, generate a quick box plot displaying the total arsenic levels by gender. Make sure that you provide appropriate labels for the x- and y- axes as well as the main title. Add some color to your plot, too! 

Save this file as a .png. Play around with widths and heights of the plots to get them just right. 
```{r setup, include=TRUE}

NHANES <- read.csv("In_Class_Exercises_Arsenic.csv", header=TRUE, sep=",")
#head(NHANES)

boxplot(Arsenic_Total ~ Gender, data = NHANES,
        xlab = "Gender",
        ylab = "Total Arsenic Levels",
        main = "Total Arsenic Levels vs Gender",
        border = c("tomato", "blue"),
        ylim = c(0, 200))

```

2) Using `ggplot2()`, generate another box plot using the data frame of NHANES data from the previous question, but this time make it a box plot grouped by gender on the y-axis. Break up each of the groups by marital status. Add jitter to each of the box plots. 

```{r}

#head(NHANES)

#summary(NHANES)  # Arsenic_Total has 218 NA's

ggplot(NHANES, aes(x = Arsenic_Total, y = Gender, fill = Marital_Status)) +
  geom_boxplot(width = 0.6,
               position = position_dodge(width = 0.75),
               outlier.shape = NA,
               alpha = 0.7) +  
  geom_jitter(aes(color = Marital_Status),
              position = position_dodge(width = 0.75),
              size = 1.5, alpha = 0.6) +  
  theme_minimal() +
  labs(title = "Arsenic Total by Gender")

```

3) Generate a violin plot of the total arsenic levels by marital status. Conduct a Wilcoxon Rank Sum test and report the p-value between the Married and Never Married participants on the violin plots. 

```{r}

library(ggpubr)

#p <- ggplot(gene_metadata, aes(x = timepoint, y = CXCL13, fill = timepoint)) +
#  geom_boxplot() +
#  labs(x = "Time Point",
#       y = "CXCL13 Count",
#       fill = "Time Point", 
#       title = "CXCL13 Count Distribution by Treatment Time Point") +
#  theme_minimal() +
#  scale_fill_brewer(palette = "Set1") +
#  stat_compare_means(method = "wilcox")
#print(p)

#NHANES_Minimal <- NHANES %>%
#  filter(Marital_Status == "Never Married" | Marital_Status == "Married")

ggplot(NHANES_Minimal, aes(x = Marital_Status, y = Arsenic_Total)) +
  geom_violin() +
  labs(x = "Marital Status",
       y = "Arsenic Total",
       title = "Arsenic Total by Marital Status from NHANES Dataset") +
  theme_minimal() +
  scale_fill_brewer(palette = "Set1") +
  stat_compare_means(method = "wilcox")

```

4) Generate your own random data of 200 participants containing the following columns: Participant_ID, Age, Sex, BMI, Ferritin_Level, and Disease_Status. Make it so that the "Cases" have generally lower ferritin levels. Normal ferritin levels are ~20–500 ng/mL in males and ~20–200 ng/mL in women. Make sure that BMIs are reasonable too! BMI is usually a normally distributed statistic centered around 25 (normal range: ~18.5–30+). Age should be just adults 18-99. Having done that, provide summary statistics for the ferritin levels. 

```{r}
library(tidyverse)
#Build the same data frame with a pipe
set.seed(123)

random_data_tbl <- tibble(
  Participant_ID = str_c("Participant_", 1:200),
  Age        = pmax(18, pmin(99, round(rnorm(200, mean = 67.5, sd = 20.66)))),
  Sex        = sample(c("Male","Female"), 200, replace = TRUE),
  Disease_Status = sample(c("Case","Control"), 200, replace = TRUE),
  # Assume a perfectly symmetric interval 18.5 - 31.5.
  # The distance from the mean to either edge is 6.5
  # Because 95% of data in a normal distribution falls within 1.96 standard deviations    
  # of the mean, you calculate the standard deviation like this 6.5/1.96
  BMI        = pmax(10, pmin(40, round(rnorm(200, mean = 25, sd = 6.5/1.96))))
) %>%
  mutate(Ferritin_Level = ifelse(Sex == "Male", 
                              rnorm(n(), mean = 25, sd = 2.9), 
                              rnorm(n(), mean = 15, sd = 2.9)))

random_data_tbl%>%                     # ← pipe: take the tibble and…
  slice_head(n = 6)       # ← show the first six rows (same as `head()`)

```

5) Convert the data frame to long format from wide format on the Age and Ferritin_Level columns. Remove the Sex column from this data frame. 
```{r}

#random_data_age_sex <- pivot_longer(random_data,
#                                           cols=c("Age","Heart_Rate"))

random_data_tbl_long <- random_data_tbl %>%
    select(-Sex) %>%
    pivot_longer(.,cols=c("Age","Ferritin_Level"))

random_data_tbl_long

```



